---
title: "Lec12_StringvFactor"
author: "Andy Turner"
date: "2023-10-17"
output: github_document
---

## Strings vs. Factors

Strings: character variables without much of an order
Factor: categorical variable, usually ordered

`stringr` is a common package

Concatenating: joining snippets into a long string
Shortening, subsetting, truncating
Changing cases
Replacing one string segment with another

`forcats` (built into the tidyverse as well)

Most of what we are doing with factors is controlling the behind the scenes in R. We are controlling the order potentially to get the correct ordering. 

```{r}
library(rvest)
library(tidyverse)
library(p8105.datasets)
```

## Strings and manipulations

```{r}
string_vec= c("my", "name", "is", "andy")

str_detect(string_vec, "andy")
str_detect(string_vec, "y")

str_replace(string_vec, "andy", "Andy")
```
str_detect: do we see this thing in the vector we are asking about? And the answer is no, no, no, yes! 
- Case sensitive as is R

str_replace: string replace - replaces the first text with the second text

```{r}
string_vec = c(
  "i think we all rule for participating",
  "i think i have been caught",
  "i think this will be quite fun actually",
  "it will be fun, i think"
  )

str_detect(string_vec, "i think")
str_detect(string_vec, "^i think")
```
`^` in this context indicates the beginning of a line. we only want "i think" that start the line!

```{r}
string_vec = c(
  "Time for a Pumpkin Spice Latte!",
  "went to the #pumpkinpatch last weekend",
  "Pumpkin Pie is obviously the best pie",
  "SMASHING PUMPKINS -- LIVE IN CONCERT!!"
  )

str_detect(string_vec,"[Pp]umpkin")
```
`[Pp]` in this context lets us look for both "Pumpkin" and "pumpkin" at the same time!

```{r}
string_vec = c(
  '7th inning stretch',
  '1st half soon to begin. Texas won the toss.',
  'she is 5 feet 4 inches tall',
  '3AM - cant sleep :('
  )

str_detect(string_vec, "[0-9]") #1
str_detect(string_vec, "^[0-9]") #2
str_detect(string_vec, "^[0-9][a-z]") #3
str_detect(string_vec, "^[0-9][a-zA-Z]) #4
        
```
1. Only matching ones with a number
2. Starts with a number
3. Starts with number then followed by lower case
4. Starts with number then followed by any letter

```{r}
string_vec = c(
  'Its 7:11 in the evening',
  'want to go to 7-11?',
  'my flight is AA711',
  'NetBios: scanning ip 203.167.114.66'
  )

str_detect(string_vec, "7.11")
```
`.` in this context says anything, so the phrase above says match anything that has 7 then 11 with anything between the two numbers. Anything that is one digit

```{r}
string_vec = c(
  'The CI is [2, 5]',
  ':-]',
  ':-[',
  'I found the answer on pages [6-7]'
  )

str_detect(string_vec, "\\[")
```
If we wanted to find a `[` we need to designate it as a special character to make it clear that we want to find the square bracket. We need to use a `\` to show it is a special character and then because `\` itself is a special character we need another `\`. Therefore, format for special character is `\\[`

"\\[[0-9]" means that we want to find any string that has [# within the string. 

## Factors

```{r}
vec_sex= c("male", "male", "female", "female")
vec_sex_f= factor(c("male", "male", "female", "female"))

as.numeric(vec_sex_f)

vec_sex_f= fct_relevel(vec_sex_f, "male")
vec_sex_f
```
`fct_relevel` our way to relevel the orders of the factors, seems harder if you have more than 2 levels. 

All factor functions start with `fct_`

```{r}
nsduh_url = "http://samhda.s3-us-gov-west-1.amazonaws.com/s3fs-public/field-uploads/2k15StateFiles/NSDUHsaeShortTermCHG2015.htm"

table_marj = 
  read_html(nsduh_url) |> 
  html_table() |> 
  first() |>
  slice(-1)
```

**Cleaning up data**
```{r}
marj_df = 
  table_marj |>
  select(-contains("P Value")) |>
  pivot_longer(
    -State,
    names_to = "age_year", 
    values_to = "percent") |>
  separate(age_year, into = c("age", "year"), sep = "\\(") |>
  mutate(
    year = str_replace(year, "\\)", ""),
    percent = str_replace(percent, "[a-c]$", ""),
    percent = as.numeric(percent)) |>
  filter(!(State %in% c("Total U.S.", "Northeast", "Midwest", "South", "West")))
```

```{r}
marj_df |>
  filter(age == "12-17") |> 
  mutate(State = fct_reorder(State, percent)) |> 
  ggplot(aes(x = State, y = percent, color = year)) + 
    geom_point() + 
    theme(axis.text.x = element_text(angle = 90, hjust = 1))

marj_df |>
  filter(age == "12-17") |> 
  mutate(State = fct_reorder(State, percent, .desc=TRUE)) |> 
  ggplot(aes(x = State, y = percent, color = year)) + 
    geom_point() + 
    theme(axis.text.x = element_text(angle = 90, hjust = 1))
```
`fct_reorder` to get our States re-ordered by what we are mapping. In this case we are mapping percent, so we wanted it to be reordered by that

`.desc=TRUE` allows me to reorder the opposite way

## NYC Restaurant Inspections

```{r}
data("rest_inspec")

rest_inspec |> 
  count(boro, grade) |> 
  pivot_wider(
    names_from=grade,
    values_from= n
  ) |> 
  knitr::kable()
```

```{r}
inspection_df=
  rest_inspec |> 
  filter(grade %in% c("A", "B", "C"), boro != "Missing") |> 
  mutate(
    boro= str_to_title(boro),
    dba= str_to_upper(dba))
```

Pizza places - pizza pizza (only those with pizza in title)
```{r}
inspection_df |> 
  filter(str_detect(dba, "PIZZA")) |> 
  count(boro)
```

```{r}
inspection_df |> 
  filter(str_detect(dba, "PIZZA")) |> 
  mutate(boro= fct_infreq(boro)) |> 
  ggplot(aes(x=boro, fill=grade))+
  geom_bar()
```
`fct_infreq` reorders the factor by frequency. 

```{r}
inspection_df |> 
  filter(str_detect(dba, "PIZZA")) |> 
  mutate(
    boro = fct_infreq(boro),
    boro = str_replace(boro, "Manhattan", "The City"))|> 
  ggplot(aes(x=boro, fill=grade))+
  geom_bar()
```
Screws up the ordering above, also string replace made it weird since it is now a factor sort of

```{r}
inspection_df |> 
  filter(str_detect(dba, "PIZZA")) |> 
  mutate(
    boro = fct_infreq(boro),
    boro = replace(boro, which(boro == "Manhattan"), "The City"))|> 
  ggplot(aes(x=boro, fill=grade))+
  geom_bar()
```

```{r}
inspection_df |> 
  filter(str_detect(dba, "PIZZA")) |> 
  mutate(
    boro = fct_infreq(boro),
    boro = fct_recode(boro, "The City" = "Manhattan"))|> 
  ggplot(aes(x=boro, fill=grade))+
  geom_bar()
```

